An efficient feedback calibration algorithm for direct imaging radio telescopes

### 2016-02-25 cal_demo/ms.tex
Starting to address Nithyas comments on cal paper

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the need for a real-time calibration method for direct imaging radio astronomy correlators. Direct imaging involves a spatial fast Fourier transform of antenna voltages, alleviating the harsh $\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 \Nant)$, which can save orders of magnitude in computation cost for next generation arrays consisting of hundreds to thousands of antennas. However, because signals are mixed in the correlator, gain correction must be applied on the front end. We develop the EPICal algorithm to form gain solutions in real time without ever forming visibilities. This method scales as the number of antennas, and produces results comparable to those from visibilities. Through simulations and application to Long Wavelength Array data we show this algorithm is a feasible solution for next generation instruments.
\end{abstract}

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. Traditionally, pair-wise visibilities are written to disk and used to calibrate offline. However, a direct imaging correlator mixes the signals from all antennas before averaging and writing to disk, making calibration a requirement at the front end ({\bf need to make this distinction more clear by pointing where exactly calibration happens in a X-based system and MOFF}). Previous solutions have involved applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

The MOFF algorithm next calls for a calibration. We will assume we have formed an estimate of the gains after $n$ iterations of a calibration loop, and derive an updated {\bf estimate for the ($n+1$)-{\it th} iteration}. How the calibration gain is applied can depend on the dominant source of the noise in the measurement. If the noise is dominated by sky fluctuations, the noise level is affected by the antenna gains so dividing by the complex gains will flatten the noise across antennas. However, if the noise is dominated in the receiver (after the complex gain factor of equation~\ref{eq:apply_gain}), a lower amplitude gain results in lower signal to noise for that antenna. Therefore, one should multiply by the conjugate of the gain to down weight {\bf it} appropriately while removing the phase acquired. We will abstain from this choice for now, and instead use a multiplicative factor $h^{(n)}_a$ to represent the application of our $n^{\mathrm{th}}$-loop estimate of the gain for antenna $a$:
\begin{equation}
\Erest{a} = h^{(n)}_a \Er{a}
\end{equation}
where,
\begin{equation}
h^{(n)}_a=\begin{cases}
1/g^{(n)}_a, & \mbox{sky noise dominated} \\ 
g^{*(n)}_a, & \mbox{receiver noise dominated}.
\end{cases}
\end{equation}

@@

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, we find,
\begin{align}\label{eq:cna}
\Cna & = \left<g^T_a \Er{a}^T(t) \frac{1}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b\Er{b}^{T*}(t) e^{-2\pi i \spix \cdot \rb} \beamtheta^*_b(\spix)\right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left<\Er{a}^T \Er{b}^{*T} \right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V^T_{ab}
\end{align}
where in the second step we group time-dependent terms, and in the third we define the true visibilities as the correlation between true antenna electric field measurements. It is easy to see from here that the net effect of including the receiver noise term, $\widetilde{n}_a(f,t)$, will result in added noise on the true visibilities, including a bias on the auto-correlation terms. In principle a bias can result on any visibility from any noise correlations between antennas, but the implementation included in the EPIC software package restricts the noise term to include only the auto-correlation noise bias, and assumes baseline dependent noise {\bf has} zero mean.

@@

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the difference is a longer effective integration for the damped version (lower thermal noise) ({\bf the last sentence is too condensed that it is not clear what you are trying to say}).

We show schematically the process of calibrating a direct imaging correlator in figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-engine and correlated against the output image pixel of interest ({\bf i.e., equation~\ref{eq:Cna} is ``on-chip''}). The correlated values are then passed off-chip to estimate the gains using equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back to the correlator to update the calibration for subsequent data streams.

@@

An important feature to note is that, like the MOFF-generated images themselves, equations~\ref{eq:cal_solution} and~\ref{eq:cal_solution_simple} include the antenna auto-correlations (the sum is over \emph{all} $b$, not excluding $a$). It can be difficult to perfectly model the noise bias from auto-correlations, which can often times be far brighter than the visibilities themselves. It can therefore be beneficial to subtract this term directly from \Cna, and exclude the $b=a$ term in the sum.
\begin{equation}
\Cna \rightarrow \Cna - \frac{1}{\Nant} h^{*(n)}_a\beamtheta^*_a e^{-2\pi i \spix \cdot \ra} \left<|\Er{a}|^2\right>_t
\end{equation}
This requires generating these correlations, which again only scale as $\mathcal{O}(\Nant)$, and are generally useful for array diagnostics. ({\bf What happened to the $g_a^T$ term outside the sum? And why did you drop $\hat{\mathbf{s}}_0$ in $W_a^*$}?)

@@

For our unknown gains, we create a set of random numbers with amplitude $1\pm0.25$ ({\bf replace with amplitude in range [0.75, 1.25]}), and completely random phase. These are our ``true gains", and we apply them to the frequency-domain simulated antenna electric fields as in equation~\ref{eq:apply_gain}. Our analysis is blind to these values until the end of the process to check accuracy. The gain estimates are initialized with unity, $g^{(0)}_a=1$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes converge toward their true values around the same time as the phases (iteration $\sim$15). At the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its true value (red line). These unbalanced high estimates caused the entire set of gains to be under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced amplitudes have been damped out and the calibration continues. Without the damping factor, the oscillation seen in the first couple iterations would have been significantly larger and taken much longer to fade out. ({\bf Add a sentence or two to explain the oscillation after convergence due to intrinsic source fluctuations.})

@@

Images created at the beginning of calibration and at the end are shown in figure~\ref{fig:sim_images}. Each image is {\bf obtained over} 10~ms integration, corresponding to all snapshot images created with a given set of gain estimates. The top panel shows the image produced with our initialized unity gains. Because the phases are completely random, the image is essentially noise with the primary beam evident. After 20 iterations, the image is far more clear, shown in the bottom panel. Each of the ten simulated sources are clearly visible, indicated with red circles.

@@

\section{Application to LWA data}\label{sec:data}
We next demonstrate our calibration algorithm using an observation from the LWA station in New Mexico. The data is from the LWA narrow-band transient buffer (TBN), with time ordered voltage data from 255 antennas within a core radius of 100~m. The central frequency is 74.03 MHz, {\bf with} a bandwidth of 100 kHz and writeout timescale of 5.12 ms (frequency channel resolution of 195.3125 Hz). For this demonstration we limit ourselves to a single polarization.

After correcting for geometric cable delays, the instrument is naturally well calibrated, as was seen in the demonstration of the EPIC imager in \citealt{thy15c}. However, we will aim to improve {\bf further} on this calibration using our algorithm.

@@

We made several choices while studying the behavior of the LWA data to improve our calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated images was consistent with average gain amplitudes of 0.25. To allow the calibration to converge quickly we initialized our gain estimates at this level. We also found a boost in signal to noise is achieved easily by averaging frequency channels and assuming the gains are constant within a sub-band. Here we average solutions across 150 channels, or about 29 kHz. With a fractional bandwidth $B/f_0 = 3.9 \times 10^{-4} \ll 1$, {\bf we assume a smooth bandpass across the band}. % Because there is still a fair amount of noise in the data, 
A damping factor of $\gamma = 0.7$ was adopted.

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, and 244 were flagged.} produced unstable gain solutions, and in fact corrupted the ensemble ({\bf specifically what is the ensemble exactly?}). We therefore flagged these antennas {\bf in} our analysis, resulting in a total of 248 antennas to calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a total of 1.536 seconds. The results of this {\bf calibration} experiment are shown in figures~\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent near the center of the image, and Cas~A is also clearly visible in the upper right. The middle panel shows the image produced after calibration with identical integration time and bandwidth. The sidelobes ({\bf or is it better to call them speckles due to uncalibrated phases because sidelobes are typically associated with the psf and not with calibration?}) throughout the image are significantly suppressed, and the galactic plane is much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux from the global sky model (GSM, \citealt{deo08}). We show the GSM modulated by two factors of the LWA {\bf power pattern} in the right panel of figure~\ref{fig:data_images} for reference.

@@

To compare the image qualities quantitatively we compute the dynamic range, defined as the peak of the image over the noise level. We estimate the noise level as the median of the absolute deviation of the image, {\bf a robust statistic}. With this metric we find the calibrated image to have a 55\% dynamic range improvement over the uncalibrated image.

@@

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 {\bf frequency channels each of width 40~kHz}. Because our calibration loop treats frequency channels independently, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.

>>>>

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the need for a real-time calibration method for direct imaging radio astronomy correlators. Direct imaging involves a spatial fast Fourier transform of antenna voltages, alleviating the harsh $\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 \Nant)$, which can save orders of magnitude in computation cost for next generation arrays consisting of hundreds to thousands of antennas. However, because signals are mixed in the correlator, gain correction must be applied on the front end. We develop the EPICal algorithm to form gain solutions in real time without ever forming visibilities. This method scales as the number of antennas, and produces results comparable to those from visibilities. Through simulations and application to Long Wavelength Array data we show this algorithm is a promising solution for next generation instruments.
\end{abstract}

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a traditional FX or XF correlator, pair-wise visibilities are written to disk and used to calibrate offline before further processing such as imaging. However, a direct imaging correlator mixes the signals from all antennas before averaging and writing to disk, making calibration a requirement at the front end, before imaging and averaging. Previous solutions have involved applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

The MOFF algorithm next calls for a calibration. We will assume we have formed an estimate of the gains after $n$ iterations of a calibration loop, and derive an updated estimate for the ($n+1$)$^{\mathrm{th}}$ iteration. How the calibration gain is applied can depend on the dominant source of the noise in the measurement. If the noise is dominated by sky fluctuations, the noise level is affected by the antenna gains so dividing by the complex gains will flatten the noise across antennas. However, if the noise is dominated in the receiver (after the complex gain factor of equation~\ref{eq:apply_gain}), a lower amplitude gain results in lower signal to noise for that antenna. Therefore, one should multiply by the conjugate of the gain to down weight {\bf it} appropriately while removing the phase acquired. We will abstain from this choice for now, and instead use a multiplicative factor $h^{(n)}_a$ to represent the application of our $n^{\mathrm{th}}$-loop estimate of the gain for antenna $a$:
\begin{equation}
\Erest{a} = h^{(n)}_a \Er{a}
\end{equation}
where,
\begin{equation}
h^{(n)}_a=\begin{cases}
1/g^{(n)}_a, & \mbox{sky noise dominated} \\ 
g^{*(n)}_a, & \mbox{receiver noise dominated}.
\end{cases}
\end{equation}

@@

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, we find,
\begin{align}\label{eq:cna}
\Cna & = \left<g^T_a \Er{a}^T(t) \frac{1}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b\Er{b}^{T*}(t) e^{-2\pi i \spix \cdot \rb} \beamtheta^*_b(\spix)\right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left<\Er{a}^T \Er{b}^{*T} \right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V^T_{ab}
\end{align}
where in the second step we group time-dependent terms, and in the third we define the true visibilities as the correlation between true antenna electric field measurements. It is easy to see from here that the net effect of including the receiver noise term, $\widetilde{n}_a(f,t)$, will result in added noise on the true visibilities, including a bias on the auto-correlation terms. In principle a bias can result on any visibility from any noise correlations between antennas, but the implementation included in the EPIC software package restricts the noise term to include only the auto-correlation noise bias, and assumes baseline dependent noise has zero mean.

@@

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the damped version is essentially a weighted average over the past several iterations, giving it a longer effective integration time.

We show schematically the process of calibrating a direct imaging correlator in figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is ``on-chip''). The correlated values are then passed off-chip to estimate the gains using equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back to the correlator to update the calibration for subsequent data streams.

@@

An important feature to note is that, like the MOFF-generated images themselves, equations~\ref{eq:cal_solution} and~\ref{eq:cal_solution_simple} include the antenna auto-correlations (the sum is over \emph{all} $b$, not excluding $a$). It can be difficult to perfectly model the noise bias from auto-correlations, which can often times be far brighter than the visibilities themselves. It can therefore be beneficial to subtract this term directly from \Cna, and exclude the $b=a$ term in the sum.
\begin{equation}
\Cna \rightarrow \Cna - \frac{1}{\Nant} h^{*(n)}_a\beamtheta^*_a(\spix) e^{-2\pi i \spix \cdot \ra} \left<|\Er{a}|^2\right>_t
\end{equation}
This requires generating these correlations, which again only scale as $\mathcal{O}(\Nant)$, and are generally useful for array diagnostics. ({\bf What happened to the $g_a^T$ term outside the sum? And why did you drop $\hat{\mathbf{s}}_0$ in $W_a^*$}?)

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is drawn from approximately gaussian distribution centered around 1 with width 0.25, and completely random phase. These are our ``true gains", and we apply them to the frequency-domain simulated antenna electric fields as in equation~\ref{eq:apply_gain}. Our analysis is blind to these values until the end of the process to check accuracy. The gain estimates are initialized with unity, $g^{(0)}_a=1$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes converge toward their true values around the same time as the phases (iteration $\sim$15). At the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its true value (red line). These unbalanced high estimates caused the entire set of gains to be under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced amplitudes have been damped out and the calibration continues. Without the damping factor, the oscillation seen in the first couple iterations would have been significantly larger and taken much longer to fade out.

@@

After the gains have converged, we see both the phases and amplitudes to continue to fluctuate coherently. This is due to the stochastic fluctuations of the simulated sources themselves, and is limited by the sky noise. For this simulation we restricted each calibration iteration to only 10~ms integration time. In principle a calibration implementation could use short integration times to allow the gains to converge, then increase the integration time to reduce this noise.

Images created at the beginning of calibration and at the end are shown in figure~\ref{fig:sim_images}. Each image is obtained over 10~ms integration, corresponding to all snapshot images created with a given set of gain estimates. The top panel shows the image produced with our initialized unity gains. Because the phases are completely random, the image is essentially noise with the primary beam evident. After 20 iterations, the image is far more clear, shown in the bottom panel. Each of the ten simulated sources are clearly visible, indicated with red circles.

@@

\section{Application to LWA data}\label{sec:data}
We next demonstrate our calibration algorithm using an observation from the LWA station in New Mexico. The data is from the LWA narrow-band transient buffer (TBN), with time ordered voltage data from 255 antennas within a core radius of 100~m. The central frequency is 74.03 MHz, with a bandwidth of 100 kHz and writeout timescale of 5.12 ms (frequency channel resolution of 195.3125 Hz). For this demonstration we limit ourselves to a single polarization.

After correcting for geometric cable delays, the instrument is naturally well calibrated, as was seen in the demonstration of the EPIC imager in \citealt{thy15c}. However, we will aim to further improve on this calibration using our algorithm.

@@

We made several choices while studying the behavior of the LWA data to improve our calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated images was consistent with average gain amplitudes of 0.25. To allow the calibration to converge quickly we initialized our gain estimates at this level. We also found a boost in signal to noise is achieved easily by averaging frequency channels and assuming the gains are constant within a sub-band. Here we average solutions across 150 channels, or about 29 kHz. With a fractional bandwidth $B/f_0 = 3.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass across the band. 
A damping factor of $\gamma = 0.7$ was adopted.

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, and 244 were flagged.} produced unstable gain solutions, and in fact corrupted the entire array. We therefore flagged these antennas in our analysis, resulting in a total of 248 antennas to calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a total of 1.536 seconds. The results of this calibration experiment are shown in figures~\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

Figure~\ref{fig:data_images} shows the improvement in the images due to our calibration. The left panel shows the uncalibrated image integrated over 51.2 ms, 29 kHz. Cyg~A is prominent near the center of the image, and Cas~A is also clearly visible in the upper right. The middle panel shows the image produced after calibration with identical integration time and bandwidth. The sidelobes ({\bf or is it better to call them speckles due to uncalibrated phases because sidelobes are typically associated with the psf and not with calibration?}) throughout the image are significantly suppressed, and the galactic plane is much more evident, despite only modeling Cyg~A and Cas~A. We also note that the feature just to the right of Cyg~A is dimmer in the calibrated image, better matching the expected flux from the global sky model (GSM, \citealt{deo08}). We show the GSM modulated by two factors of the LWA power pattern in the right panel of figure~\ref{fig:data_images} for reference.

@@

To compare the image qualities quantitatively we compute the dynamic range, defined as the peak of the image over the noise level. We estimate the noise level as the median of the absolute deviation of the image, a robust statistic. With this metric we find the calibrated image to have a 55\% dynamic range improvement over the uncalibrated image.

@@

We use a simulated sky consisting of a 5 Jy calibrator source, and 10 other random sources with apparent flux densities 0.2 to 0.5 Jy (total sky power $\approx$~8 Jy). We generated these sources randomly, but kept them fixed for each run. As in section~\ref{sec:sim}, we use the MWA core layout (51 antennas), and a 4.4~m square tophat antenna voltage pattern. We simulate 64 frequency channels each of width 40~kHz. Because our calibration loop treats frequency channels independently, each channel can be treated as a separate trial of the simulation, and is used to better estimate the statistics.
